Management induced changes of soil organic carbon on global croplands

### 2021-08-17 SOCbudgetPaper/methods.Rmd
update method part and pdf

# Methods

@@

We calculate $SOC$ stocks based on the area shares of land-use types ($lu$) within our grid cells ($i$). If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated within our budget. As we do not distinguish between freshly converted and old cropland (see \@ref(fig:lutransitions)), we update the average SOC stock per area within this modeling step. We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

with $A_{lu}$ being the land-use type specific areas, $AR_{lu}$ and $AE_{lu}$ the area reduction resp. area expansion of the two land-use types. Data sources and methodology on land-use states and changes are described in Sect. \@ref(sec:landuse).

@@

with $T_{c,i}$ being the translation matrix for grid cells $i$ into corresponding climate zones $c$. For this analysis, we use the default $F^{\mathrm{SCF}}$ from the Tier 1 method of Eggleston et al. [-@eggleston_ipcc_2006] and Calvo Buendia et al. [@calvo_buendia_ipcc_2019] as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Crop residue production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and key aspects are explained here given its central role for soil carbon modeling. Starting from harvested crop production ($CP$) estimates and their respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific harvest index values ($HI$) and root:shoot ratios ($RS$) as follows

\begin{equation}
\begin{aligned}
AGR_{i,t,cg} & = CP_{i,t,cg} \cdot HI^{\mathrm{prod}}_{cg} + CA_{i,t,cg} \cdot HI^{\mathrm{area}}_{cg}
\qquad & \textrm{and} \\
BGR_{i,t,cg} & = (CP_{i,t,cg} + AGR_{i,t,cg}) \cdot RS_{cg} \qquad                                            & \forall\quad cg, i, t.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a production ($HI^{\mathrm{prod}}$) and an area dependent ($HI^{\mathrm{area}}$) fraction [@eggleston_ipcc_2006]. Deviating from Bodirsky et al. [-@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

@@

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for these three drivers: In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set). The \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in $\unit{t ha^{-1}}$) at the level of 1975 onwards. Within the \textit{constResidue} scenario at different effects overlay each other: yields and with them residue biomass increase due to productivity gains; rates of residue left or returned to fields are raising; and shifts of cropping pattern change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios \textit{constTillage}, \textit{constResidues} and \textit{constManure}.

### Initialization choice {#sec:scen_initial}

As outlined in Sect. \@ref(sec:initsoc) we assume landuse-type specific steady-state SOC stocks for the start year 1901 followed by a long spin-up period of 74 years (\textit{Initial-lu}). As some SOC compartments decay over very long timescales, the initialisation setting might strongly effect the overall outcome of SOC stocks and changes. Thus we conduct counterfactual scenario \textit{Initial-natveg}, initializing the start year with the steady-state SOC under potential natural vegetation for all land-use types. We additionally combined the \textit{Initial-natveg} scenario with the \textit{constManagement} scenario.

### Parameterization of natural litter fall {#sec:scenlitterpnv}

Due to missing information about the litterfall composition, we assume woody potential natural vegetation for the lignin-to-carbon (short LG:C) and nitrogen-to-carbon (short N:C) ratios for all land under natural vegetation. This is however a strong simplification. Therefore, we test the sensitivity of the model for other natural vegetation parameterizations. Perennial grasses are given by the IPCC guidelines [@calvo_buendia_ipcc_2019], but for woody biomass we used the CENTURY configuration file [@century_model_2000] for trees as parameterization were not given. Whereas LG:C ratios ranges between $0.09-0.35$ with an average of $0.24$ (default value) for different tree compartments and different tree types, N:C ratios range between $0.001-0.05$ with an average of $0.015$ (default value). We decided to average over all tree compartments and tree types equally for the default values, since we have no data available on tree type distribution and tree compartment composition. We test however a range of different plant types as specified in Table \@ref(tab:scenlitterpnv).

 \begin{table*}[h]
 \caption{Scenarios for sensitivity analysis on litter parameterization}
 \begin{tabular}{l l l l}
 \tophline
  Scenario name & plant type & N:C & LG:C \\
  \middlehline
  \textit{LitterPNV-Mean} & mix over all tree types & 0.015  & 0.25 \\
  \textit{LitterPNV-Grass} & perennial grasses     & 0.029  & 0.11 \\
  \textit{LitterPNV-AASS}  & Arctic alpine subshrub           & 0.022  & 0.22 \\
  \textit{LitterPNV-BLEAF} & Temperate broadleaf evergreen    & 0.013  & 0.27 \\
  \textit{LitterPNV-BRCF}  & Boreal coniferous forest         & 0.013  & 0.17 \\
  \textit{LitterPNV-MTCF}  & Maritime coniferous forest       & 0.0094 & 0.22 \\
  \textit{LitterPNV-TREF}  & Tropical evergreen forest        & 0.016  & 0.30 \\
 \bottomhline
 \end{tabular}
 \belowtable{}
 (\#tab:scenlitterpnv)
 \end{table*}
```

>>>>

---
output:
  pdf_document: default
  html_document: default
---
# Methods

@@

```{r captionLUtrans, echo=FALSE}
captionLUtrans <- "Scheme of land-use transition representation for a simplified case.."
```

```{r LUtrans, out.width = "16cm", fig.cap = captionLUtrans, echo=FALSE, fig.pos="h"}
# All defaults
knitr::include_graphics("../ResultNotebooks/Output/Images//LUtransition.png")
```

We calculate $SOC$ stocks based on the area shares of land-use types ($lu$) within our grid cells ($i$). If land is converted from one land-use type $lu=\{crop,natveg\}$ into the other $!lu=\{natveg,crop\}$, a respective share of the $SOC$ is reallocated within our budget. As we do not distinguish between newly converted and old cropland (see \@ref(fig:lutransitions)), we update the average SOC stock per area within this modeling step. We account for land conversion at the beginning of each time step $t$ by calculating a preliminary stock $SOC_{t^*}$ via

@@

with $A_{lu}$ being the land-use type specific areas, $AR_{lu}$ the area reduction and $AE_{lu}$ the area expansion of the two land-use types. Data sources and methodology on land-use states and changes are described in Sect. \@ref(sec:landuse).

@@

with $T_{c,i}$ being the translation matrix for grid cells $i$ into corresponding climate zones $c$. For this analysis, we use the default $F^{\mathrm{SCF}}$ from the Tier 1 method of Eggleston et al. [-@eggleston_ipcc_2006] and Calvo Buendia et al. [-@calvo_buendia_ipcc_2019] as a comparison and consistency check for our more detailed Tier 2 steady-state approach.

@@

Crop residue production and management is based on a revised methodology of Bodirsky et al. [-@bodirsky_n2o_2012] and key aspects are explained here given its central role for soil carbon modeling. Starting from crop yield estimates ($Y$) and respective harvested crop area ($CA$), we estimate above-ground ($AGR$) and below-ground ($BGR$) residual biomass using crop group ($cg$) specific ratios for above-ground residues to harvested biomass ($r^{\mathrm{ag,prod}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{tC}}$), above-ground residues to harvested area ($r^{\mathrm{ag,area}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{ha}}$) and below-ground residues to above-ground biomass ($r^{\mathrm{bg}}_{cg}$ in $\tfrac{\unit{tC}}{\unit{tC}}$) as follows

\begin{equation}
\begin{aligned}
AGR_{i,t,cg} & = CA_{i,t,cg} \cdot \left( Y_{i,t,cg} \cdot  r^{\mathrm{ag,prod}}_{cg}+  r^{\mathrm{ag,area}}_{cg} \right)
\qquad & \textrm{and} \\
BGR_{i,t,cg} & = (CP_{i,t,cg} + AGR_{i,t,cg}) \cdot r^{\mathrm{bg}}_{cg} \qquad                                            & \forall\quad cg, i, t.
(\#eq:resbiomass)
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the above-ground residue calculations into a production ($r^{\mathrm{ag,prod}}$) and an area dependent ($r^{\mathrm{ag,area}}$) fraction [@hergoualch_n2o_in_ipcc_2019]. Deviating from Bodirsky et al. [-@bodirsky_n2o_2012] we use harvested instead of physical crop area to account for increased residue biomass due to multiple cropping and decreased residue amounts due to fallow land. We assume that all $BGR$ are left in the soil, whereas $AGR$ can be burned or harvested for other purposes such as feeding animals [@weindl_livestock_2017], fuel or for material use.

@@

To single out the impact of tillage practices, residue and manure inputs, we defined scenario with constant values for these three drivers: In the \textit{constTillage} scenario the adoption of no-tillage practices are neglected (adoption starts in 1974 according to the available data set). The \textit{constResidues} and the \textit{constManure} scenario assume constant input rates from residues resp. manure (in $\unit{t ha^{-1}}$) at the level of 1975 onward. Within the \textit{constResidue} scenario at different effects overlay each other: yields and with them residue biomass increase due to productivity gains; rates of residue left or returned to fields are raising; and shifts of cropping pattern change the amount of residue biomass due to crop-group specific harvest index values. The \textit{constManagement} combines all three scenarios \textit{constTillage}, \textit{constResidues} and \textit{constManure}.
